---
title: "Homework 9 (ISYE 6501) - Question 12.1 & 12.2"
output:
  pdf_document:
    latex_engine: xelatex
---
Sorry in advance to my peer reviewers! (ᵕ—ᴗ—) I like to write out my reasoning and can be... wordy... But I will try to be even more brief this time after taking the advice from HW7 peers!  
<!-- ## rmarkdown::render("edX/GTx.ISYE6501/homework/homework9_answers.Rmd") -->  

## Question 12.1 Answer
I used https://www.overleaf.com/learn and AI to help me shape the code for this question, as I am not as experienced with writing R yet.  
I did not copy any specific lines of code, I just used it as a helpful reference for me to better understand how best to write it.   

A design of experiments approach for something relevant to my life would be appropriate for improving response rates to outreach emails at work.  
For example, suppose a team wants to improve the percentage of members who respond to a renewal reminder email. Instead of changing only one thing at a time, a DOE approach would let the team test several factors systematically and learn both:  
1. which factors matter most on their own, and  
2. whether combinations of factors interact with each other.  

Possible factors could include:  
- whether the subject line is formal or informal,  
- whether the email is sent in the morning or afternoon,  
- whether a deadline is mentioned,  
- whether the sender is shown as a generic department or a named staff member,  
- whether a follow-up reminder is sent.  

The response variable could be something like:  

- open rate,  
- click-through rate,  
- reply rate, or  
- completed renewal rate.  

A DOE approach would be better than changing one feature at a time because multiple parts of the message may matter simultaneously. For example, mentioning a deadline might work especially well only when combined with a follow-up reminder, or a more personal sender name might matter more for one style of subject line than another. A structured experiment would help estimate those effects more efficiently and objectively.  

## Question 12.2 Answer
This question asks for a **fractional factorial design** for 10 yes/no house features, using only 16 houses instead of all possible combinations.  

I used the `FrF2` package in R, as requested. The `FrF2` function returns a design matrix with values of `1` for "include the feature" and `-1` for "do not include the feature."  

```{r q12_2_setup, message=FALSE, warning=FALSE}
rm(list = ls())

if (!requireNamespace("FrF2", quietly = TRUE)) {
  install.packages("FrF2", repos = "https://cloud.r-project.org")
}

library(FrF2)
```

```{r q12_2_design, message=FALSE, warning=FALSE}
feature_names <- c(
  "LargeYard",
  "SolarRoof",
  "FinishedBasement",
  "SmartHome",
  "Pool",
  "Garage",
  "RenovatedKitchen",
  "HomeOffice",
  "EVCharger",
  "CornerLot"
)

house_design <- FrF2(
  nruns = 16,
  nfactors = 10,
  factor.names = feature_names,
  randomize = FALSE
)

house_design
```

The table above is the fractional factorial design. Each row represents one fictitious house to show to potential buyers, and each column is one feature. A value of:  

- `1` means the feature is included,  
- `-1` means the feature is not included.  

For a cleaner version to submit as a house-feature table, I converted the output to a data frame below and added a house number.  

```{r q12_2_table, message=FALSE, warning=FALSE}
house_table <- as.data.frame(house_design)
house_table$House <- 1:nrow(house_table)
house_table <- house_table[, c("House", feature_names)]

house_table
```

If I wanted to describe the design in words, I would say that this is a 16-run two-level fractional factorial design for 10 binary factors.  
It lets the agent estimate the most important feature effects much more efficiently than testing all `2^{10} = 1024` possible house combinations.  

For readability, here is the same design translated from `1 / -1` into `Yes / No`.  

```{r q12_2_yes_no, message=FALSE, warning=FALSE}
house_table_readable <- house_table
house_table_readable[feature_names] <- lapply(
  house_table_readable[feature_names],
  function(x) ifelse(x == 1, "Yes", "No")
)

house_table_readable
```

**ANALYSIS:**  
A full factorial experiment with 10 yes/no features would require 1024 different houses, which is wayyyy too many for this survey. The 16-house fractional factorial design is a practical compromise that still spreads the feature combinations across the survey in a systematic way. That allows the real estate agent to learn which house features appear to contribute most to perceived market value without needing to show every possible combination.
